---
title: "Storm Event Economic and Human Costs in the US, 1961-2011"
author: "Gary A. Miller"
date: "Friday, August 22, 2014"
output: html_document
---

### Synopsis ###

This report describes an analysis of data from the US NOAA's 
<a href="http://www.ncdc.noaa.gov/">National Climactic Data Center</a>.  The goal
is to determine which types of storm events have caused the most damage to 
property and to human life.

### Data Processing ###

#### Loading and Pre-processing the Data ####

This code creates a folder for the data to be stored in if it does not
already exist.

```{r folder}
if(!file.exists("data")) {
dir.create("data")
}
```

This code downloads the file into the Data folder.  The Download Date is also 
recorded.

```{r download}
FileUrl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(FileUrl,destfile="./data/StormData.csv.bz2")

DownloadDate <- format(Sys.Date(),"%B %d %Y")
```

These results are effective as of `r DownloadDate`.

The next step is to read in the data.

``` {r read, cache=TRUE}
StormData <- read.csv(bzfile("./data/StormData.csv.bz2"))
```

Initial exploration shows that this is a fairly large dataset, which includes 
some attributes that are not useful to our specific analysis.

``` {r explore}
dim(StormData)
names(StormData)
```

My first transformation is to change the datatype of the Date column from factor
to POSIXlt, and to extract just the year into a new column. 

```{r datetransform}
StormData$BGN_DATE <- strptime(StormData$BGN_DATE,format="%m/%d/%Y")
StormData$BGN_YEAR <- format(StormData$BGN_DATE,"%Y")
```

This analysis is unavoidably shaped by the fact that much more data exists for 
events in more recent years.  While data exists from events as far back as 1950,
I removed data from 1960 and earlier.

```{r yearsubset}
StormData <- StormData[StormData$BGN_YEAR>=1961,]
```

```{r yearplot}
library(ggplot2)
library(scales)
    
h <- ggplot(StormData,aes(x=BGN_YEAR))
h2 <- h + geom_histogram(fill="blue") +
    scale_x_discrete(breaks=seq(1961,2011,5)) +
     labs(x="Year",y="Number of Storms",
          title="Number of Storms by Year, 1961 - 2011")

print(h2)
```

The table below shows the number of storms identified as having taken place in
each state.  I removed those not placed in one of the fifty US states or in the
District of Columbia.

```{r statesubset}
table(StormData$STATE)

States <- c("AK","AL","AR","AZ","CA","CO","CT","DC","DE","FL","GA","HI","IA",
            "ID","IL","IN","KS","KY","LA","MA","MD","ME","MI","MN","MO","MS",
            "MT","NC","ND","NE","NH","NJ","NM","NV","NY","OH","OK","OR","PA",
            "RI","SC","SD","TN","TX","UT","VA","VT","WA","WI","WV","WY")

StormData <- StormData[StormData$STATE %in% States,]
```

Many variables in this dataset are not needed to answer these specific 
questions.  I removed unnecessary columns.

```{r columnsubset}
KeepCols <- c("EVTYPE","BGN_YEAR","FATALITIES","INJURIES","PROPDMG",
              "PROPDMGEXP","CROPDMG","CROPDMGEXP")

StormData <- StormData[,names(StormData) %in% KeepCols]
```

The titles given to each event (EVTYPE) have not been applied consistently. 
The next set of code creates a new EVLABEL column, sets 'other' as the default,
and then assigns names based on a series of regular expression searches.

```{r eventtype}
InitEventTypeCount <- length(unique(StormData$EVTYPE))

StormData$EVTYPE <- toupper(StormData$EVTYPE)

StormData$EVLABEL <- "OTHER"
StormData$EVLABEL[grepl("FIRE",StormData$EVTYPE)] <- "FIRE"
StormData$EVLABEL[grepl("DUST",StormData$EVTYPE)] <- "DUST"
StormData$EVLABEL[grepl("FLOOD|SURGE",StormData$EVTYPE)] <- "FLOOD"
StormData$EVLABEL[grepl("HAIL",StormData$EVTYPE)] <- "HAIL"
StormData$EVLABEL[grepl("WIND",StormData$EVTYPE)] <- "HIGH WIND"
StormData$EVLABEL[grepl("WARM|HOT|DROUGHT|DRY|DRI|HEAT|HIGH",
                        StormData$EVTYPE)] <- "HEAT"
StormData$EVLABEL[grepl("COLD|FROST",StormData$EVTYPE)] <- "COLD"
StormData$EVLABEL[grepl("SNOW|BLIZZARD|WINT*MIX|WINT*STOR|ICE",
                        StormData$EVTYPE)] <- "SNOWSTORM"
StormData$EVLABEL[grepl("TORN|SPOUT|WALL|FUNNEL",
                        StormData$EVTYPE)]<-"TORNADO"
StormData$EVLABEL[grepl("HURRICANE|TYPHOON|TROP",
                        StormData$EVTYPE)]<-"HURRICANE"
StormData$EVLABEL[grepl("THUN|TSTM|MICROB|RAIN",
                        StormData$EVTYPE)] <- "THUNDERSTORM"

table(StormData$EVLABEL)
FinalEventLabelCount <- length(unique(StormData$EVLABEL))

```

There were initially `r InitEventTypeCount` variations of event type spellings; 
the filtering above reduces that number to `r FinalEventLabelCount`.  Although 
the number of 'Other' responses remaining indicates that this process is 
inexact.  

#### Analysis - Human Casualties ####

Combining fatality counts with injury counts yields a new column with a total 
number of casualties for each storm event.  This is the ultimate measure of 
population impact for this analysis.

```{r casualtycount}
StormData$CASUALTIES <- StormData$INJURIES + StormData$FATALITIES
```

Reshaping the data by both the fatality and injury variables allows for the 
human impact to be visualized with a stacked barplot later on in this report.

```{r casualtymelt}
library(reshape2)

StormCasualtyMelt <- melt(StormData,id="EVLABEL",measure.vars=
                        c("FATALITIES","INJURIES"))
```

#### Analysis - Property Damage ####

The documentation explains that the Property Damage Exponent corresponds to the
degree of magnitude to apply to the amount given in the Property Damage column.
I created a new 'Multiplier' column and populated it with the corresponding 
amount, with 1 as the default.  I then multiplied the multiplier by the estimate
to get a total damage amount. 

```{r PropertyDamage}

table(StormData$PROPDMGEXP)

# Populate multiplier values based on code

StormData$PROPDMGMULTIPLIER <- as.numeric(1) 

StormData$PROPDMGMULTIPLIER[StormData$PROPDMGEXP=="K"|
                            StormData$PROPDMGEXP=="k"|
                            StormData$PROPDMGEXP=="3"] <- 1000
StormData$PROPDMGMULTIPLIER[StormData$PROPDMGEXP=="4"] <- 10000
StormData$PROPDMGMULTIPLIER[StormData$PROPDMGEXP=="5"] <- 100000
StormData$PROPDMGMULTIPLIER[StormData$PROPDMGEXP=="M"|
                            StormData$PROPDMGEXP=="m"|
                            StormData$PROPDMGEXP=="6"] <- 1000000
StormData$PROPDMGMULTIPLIER[StormData$PROPDMGEXP=="7"] <- 10000000
StormData$PROPDMGMULTIPLIER[StormData$PROPDMGEXP=="8"] <- 100000000
StormData$PROPDMGMULTIPLIER[StormData$PROPDMGEXP=="B"] <- 1000000000

# Calculate total property damage

StormData$PROPDMGTOTAL <- StormData$PROPDMG * StormData$PROPDMGMULTIPLIER
```

The same process works for finding the amount of crop damage.

```{r CropExp}
table(StormData$CROPDMGEXP)

# Populate multiplier values based on code

StormData$CROPDMGMULTIPLIER <- as.numeric(1)

StormData$CROPDMGMULTIPLIER[StormData$CROPDMGEXP=="2"] <- 100
StormData$CROPDMGMULTIPLIER[StormData$CROPDMGEXP=="K"|
                            StormData$CROPDMGEXP=="k"] <- 1000
StormData$CROPDMGMULTIPLIER[StormData$CROPDMGEXP=="M"|
                            StormData$CROPDMGEXP=="m"] <- 1000000
StormData$CROPDMGMULTIPLIER[StormData$CROPDMGEXP=="B"] <- 1000000000

# Calculate total crop Damage

StormData$CROPDMGTOTAL <- StormData$CROPDMG * StormData$CROPDMGMULTIPLIER
```

Now the total amount of property damage can be calculated by adding the property
damage total to the crop damage total.

```{r rearrange}
StormData$TOTALDAMAGE <- StormData$PROPDMGTOTAL + StormData$CROPDMGTOTAL
```

Here's a look at the final dataset before aggregating.

```{r StormData}
head(StormData,5)
```

As with the casualty metric, I reshaped the data based on the totals for crop 
and property damage.

```{r meltProperty}
StormEcoMelt <- melt(StormData,id="EVLABEL",measure.vars=
                         c("CROPDMGTOTAL","PROPDMGTOTAL"))
```
### Results ###

Tornados are far and away the storm event type that cause the most damage to 
human life.

```{r plotcasualties}
c <- ggplot(StormCasualtyMelt,aes(x=EVLABEL,y=value,fill=variable))

c2 <- c + geom_bar(stat="identity") + 
    labs(x="Event Type",y="Number of Casualties") + 
    labs(title="Combined Fatalities and Injuries by Event Type") +
    scale_fill_manual(values=c("#e34a33","#fdcc8a"),name="Casualty Type") +
    scale_y_continuous(label=comma) +
    theme(axis.text.x = element_text(angle=90))

print(c2)
```

Floods have cause the most damage to property.

It's also interesting to note the comparatively high percentage of damage that
the heat / drought events have done to crops rather than other types of physical
property.

```{r plotproperty}
p <- ggplot(StormEcoMelt,aes(x=EVLABEL,y=value,fill=variable))

p2 <- p + geom_bar(stat="identity") + 
    labs(x="Event Type",y="Total Property and Crop Damage ($US)") +
    labs(title="Property Damage by Event Type") +
    scale_fill_manual(values=c("#2ca25f","#fecc5c"),name="Damage Type",
                      labels=c("CROP","PROPERTY")) + 
    scale_y_continuous(label=comma) +
    theme(axis.text.x = element_text(angle=90))

print(p2)
```

### References ###

This analysis was completed as an assignment in 
<a href="https://www.coursera.org/course/repdata">Reproducible Research</a>,
part of the Data Science Specialization offered by Johns Hopkins University via 
the Coursera platform.

The NCDC's 
<a href="https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf">Storm Center FAQ</a> and the National Weather Service's <a href="https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf">documentation</a> explain the definitions of each variable and how the data was compiled.

Additional data can be obtained from the Storm Events 
<a href="http://www.ncdc.noaa.gov/stormevents">database</a>.